Discrete Ordinates

### 2017-06-08 docprep/appdx_a.tex
Include a appendix label
Updated thesis text

%%
%% This is file `chapmin.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% ths.dtx  (with options: `chapmin')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from chapmin.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file ths.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)

 %% ... sample chapter ...

This appendix covers the math for Legendre polynomials and spherical harmonics

\section{Legendre Polynomials}

They exist.

\subsection{Spherical Harmonics}

They also exist.

The real harmonics. The even/odd harmonics.

The addition theorem.

>>>>

%%
%% Edward T. Norris
%% Discrete Ordinates Computed Tomography Organ Dose Simulator (DOCTORS)
%% 
%% === Appendix A ===
%%

This appendix gives a brief summary of flux moments and the Legendre spherical harmonics needed to compute them.

\section{Flux Moments}\label{appdx:moments}

The flux moments are an alternate representation of the angular flux  within a system. The rationale behind usage of flux moments is that they map the discrete flux ($\psi$) from a function of angle to a function of $P_N$ expansion. This reduces the amount of memory required for computation. Storing $\psi$ directly requires storing $G \times N_V \times N_a$ values where $G$ is the number of groups, $N_V$ is the number of voxels, and $N_a$ is the number of angles. Storing the flux moments, however, requires storing only $G \times N_V \times (N^2 + 2N + 1)$ values where $N$ is the $P_N$ expansion value. For example, a problem requiring $S_6$ needs $6 \times 8 = 48$ directions, but a problem requiring $P_6$ needs 49 expansion coefficients which would seem approximately eqivalent, but typical problems require far more discrete directions than expansion coefficients. Comples problems can easily require $S_{16}$ or higher but expansions above $P_5$ are rarely encountered.

\section{Legendre Polynomials}\label{appdx:leg}

The Legendre polynomials are solutions to:
\begin{equation}
\begin{split}
P_0(\mu) &= 1 \\
P_l(\mu) &= \frac{1}{2^l l!} \frac{d^l}{d\mu^l}(\mu^2 - 1)^l \,, \quad l \in \mathbb{N}
\end{split}
\end{equation}
where $\mathbb{N}$ is the set of natural numbers (1, 2, 3...). Table~\ref{tab:legendre} gives the solution of the first few Legendre polynomials. The Legendre polynomials are orthogonal on the domain $[-1, 1]$ in that
\begin{equation}
\frac{1}{2} \int_{-1}^{1} P_l(\mu) P_{l'}(\mu) d\mu = 
\begin{cases}
\frac{1}{2l+1} \,, \quad l = l'\\
0 \,, \quad \text{otherwise}
\end{cases}
\end{equation}

\begin{table}[ht]
\caption{Legendre Polynomials}
\centering 
\begin{tabular}{c c}
\hline \hline   
$P_l$    & $P_l(\mu)$ \\ [0.5ex] 
\hline
$P_0$ & 1 \\
$P_1$ & $\mu$ \\ 
$P_2$ & $\frac{1}{2}(3\mu^2 - 1)$ \\
$P_3$ & $\frac{1}{2}(5\mu^3 - 3\mu)$ \\
$P_4$ & $\frac{1}{8}(35\mu^4 - 30x^2 + 3)$ \\
$P_5$ & $\frac{1}{8}(63\mu^5 - 70\mu^3 + 15\mu)$ \\
$P_6$ & $\frac{1}{16}(231\mu^6 - 315^4 + 105^2 - 5)$ \\
$P_7$ & $\frac{1}{16}(429\mu^7 + 693\mu^5 - 315^3 + 35^3\mu)$ \\[1ex]
\hline
\end{tabular}
\label{tab:legendre}
\end{table}

The orthogonality of the Legendre polynomials allows an infinite series to exactly represent any function on the domain $[-1, 1]$ as
\begin{equation}
f(\mu) = \sum_{l=0}^{\infty} C_l P_l(\mu)
\end{equation}
with apprpriately selected constants, $C_l$ and is used to approximate functions arbitrarily
\begin{equation}
f(\mu) \approx \sum_{l=0}^{N} C_l P_l(\mu).
\end{equation}
The Legendre polynomials are extended into the associated Legendre polynomials which have an additional orthogonality.

\section{Associated Legendre Polynomials}\label{appdx:assoc}

The associated Legendre polynomials are defined as the solution to
\begin{equation}
\begin{split}
P_l^m(\mu) &= (-1)^m (1-\mu^2)^{m/2} \frac{d^m}{d\mu^m}P_l(\mu) \\
P_l^0(\mu) &= P_l(\mu)
\end{split}
\end{equation}
and exhibit the following orthogonality on $[-1, 1]$:
\begin{equation}
\frac{1}{2} \int_{-1}^{1} P_l^m(\mu) P_{}^{}(\mu) d\mu = 
\begin{cases}
\frac{1}{2l+1} \frac{(l+m)!}{(l-m)!} \,, \quad l=l' \text{ and } m=m'\\
0\,, \quad \text{otherwise}.
\end{cases}
\end{equation}
The spherical harmonics take advantage of the double orthogonality to approximate 2D distributions.

\section{Spherical Harmonics}\label{appdx:spherical}

The spherical harmonics are defined as:
\begin{equation}
Y_{lm}(\hat{\Omega}) = \sqrt{\frac{(2l+1)(l-m)!}{(l+m)!}}P_l^m(\mu)e^{e \tau m}
\end{equation}
where $\mu$ is the cosine of the angle between $\hat{Omega}$ and the $x$-axis and $\tau$ is the rotation about the $x$-axis with respect to the $y$-axis of $\hat{\Omega}$ projected onto the $yz$ plane which can be compuated as:
\begin{equation}
\tau = \frac{\eta}{\sqrt{\eta^2 + \xi^2}}.
\end{equation}
The spherical harmonics have the orghogonality
\begin{equation}
\int_{}^{}Y_{lm}(\hat{Omega})Y_{l'm'}^*(\hat{\Omega}) d\hat{\Omega}= 
\begin{cases}
1 \,, \quad l = l' \text{ and } m = m' \\
0 \,, \quad \text{otherwise}
\end{cases}
\end{equation}
where $Y_{lm}^*$ is the complex conjugate of $Y_{lm}$. Using the addition theorem which states
\begin{equation}
P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2l+1}\sum_{m=-l}^{l}Y_{lm}^*(\hat{\Omega}') Y_{lm}(\hat{\Omega})
\end{equation}
gives
\begin{equation}
P_l(\hat{\Omega} \cdot \hat{\Omega}') = P_l(\mu)P_l(\mu') + 2\sum_{m=1}^{l}\frac{(l-m)!}{(l+m)!}P_l^m(\mu) P_l^m(\mu') cos(m[\tau - \tau']).
\end{equation}

### 2017-06-08 docprep/chapbib.tex
Updated thesis text

\section{CT Dose Estimation}
As discussed in Chapter 1, the dose a patient receives from both diagnostic CT procedures and radiation therapy is of great concern to the medical community [CITE]. However, dose quantification is not currently patient specific. The air kerma (measured in Gy) measures the x-ray beam intensity and can be measured directly [CITE]. As the x-ray beam passes through the patient, it deposits energy. The energy deposition per unit mass is the absorbed dose (measured in Gy) which is then weighted by the radiation type to give the equivalent dose (measured in Sv). The equivalent dose quantifies the biological detriment to the patient from the procedure. The effective dose (measured in Sv) accounts for the radiosensitivity of each organ and is the most meaningful measure of risk to the patient. Any two procedures that result in the same effective dose to the patient carry the same acute and long-term risks. Any procedure that gives an effective dose to a patient has the same risk as any procedure that would give the same whole body dose.

@@

\citet{ref:lauk} found that severly obese patients received twice as much dose from CT procedures as others on average. Some patients received as much as five times the typical dose.

@@

Monte Carlo codes have been used in the past to compute flux-to-dose conversion factors for a reference person~\citep{ref:icrp116}. The values provided by~\citet{ref:icrp116} are energy dependent conversion factors, $H(E)$ that convert flux to effective dose for a particular orientation. The dose, $D$ is then computed
\begin{equation}
D = \int_0^\infty \varphi(E) H(E) dE
\end{equation}
where $\varphi$ is the scalar flux.

Monte Carlo codes are generally known for their accuracy; they are able to sample the problem and attain arbitrarily high precision given sufficient runtime. They also take advantage of combinatorial geometry allowing a user to produce geometrically complex 3D structures. Deterministic methods are generally much faster than their Monte Carlo counterparts but do not benefit from excess runtime. They also consume vastly more resources, particularly random-access memory (RAM). However, in regularly structures meshes, deterministic methods can greatly outperform Monte Carlo. A large number of spatial regions slows down the particle transport in Monte Carlo and increases the RAM required to store the numerous voxels.

Monte Carlo codes such as TOPAS and Geant4 \citep{ref:agostinellis} are the gold standard \citep{ref:jiax}

Codes such as EGS4, its improved versions EGSnrc, and EGS5 \citep{ref:nelsonw}, and PENELOPE \citep{ref:salvatf} are all Monte Carlo and have shown very good results to experimental. But they are developed for high energy physics.

The alternative to Monte Carlo is deterministic solutions. The best known solution techniques include the discrete ordinate method, the finite element method, and the method of characteristics. The discrete ordinate method is employed by DOCTORS and the methodology is covered independently in Section~\ref{sec:discordlit}.

One of the most well known MOC codes is DeCART. DeCART is actually a coupled solver in that it utilizes different solution modalities for the radial plane and the axial direction. It uses a more complex modality in the radial plane such as MOC and then uses diffusion to transport particles axially~\citep{ref:hursinm}.

The Method of Characteristics is a deterministic solution that has some advantages over discrete ordinates but is very complex to set up in 3D. One implementation is the TRANSMED code which was used to compute external beam therapy dose profiles accurately~\citep{ref:williamsm}.

The MOC solution has been extended to the time dependent domain by \citet{ref:hoffmana}.

Raytracing algorithms have been developed for numerous discret ordinate codes include RAY3D~\citep{ref:yingz} and ATTILLA~\citep{ref:wareingt} which utilizes unstructured tetrahedral meshes.

Dose estimations have been made using many methodologies including

COMPUTATION

@@

The reconstruction populates the CT mesh with CT numers, also known as Hounsfield units. These values are related to the attenuation coefficient of the material represented by the voxel. Traditionally, a CT number of zero corresponds to the attenuation of water while -1000 corresponds to dry air CITE though some variations do exist \citep{ref:plessisf} \citep{ref:sawc}.

@@

\section{Discrete Ordinates}\label{sec:discordlit}
The dsicrete ordinates method dates back to CITE but remains one of the most prominent solution modalities for radiation transport in use today. A comprehensive review of discrete ordinate methods is given by \citet{ref:lewise}. Additional references include [CITES].

The oldest major discrete ordinate method implementation is DORT which was superceded by its 3D counterpart TORT~\citep{ref:rhoadesw}. Since TORT, new discrete ordinate method implementations have laregly been extensions allowing more advanced computation.

@@

KBA DISCUSSION

@@

Many raytracing algorithms are based on the Bresenham line raserizing algorithm which was designed for graph plotting hardware~citep{ref:bresenhamj}. That algorithm was intende to produce lines acceptable for human interpretation and no guarantee is made that all pixels along the line will be identified making this class of algorithms inappropriate for dosimetric raytracing since attenuation through some voxels would be skipped. Some extensions have been added to it that guarantee passage through all voxels though~\citep{ref:liuy}.

\citet{ref:clearyj} proposed a voxel traversal algorithm in which they initially identify the voxel containing the starting point of the ray, and transport it to the first surface of its bounding voxel it would cross. This is done by computing the distance traveled in order to cross each enclosing surface. The surface with the least distance is the first surface crossed. Therefore, the particle is advanced to the intersection point and considered inside the next cell, having crossed the surface. This process is repeated until a termination criteria is met.

This work uses the raytracing algorithm proposed by~\citet{ref:wooa} which is very similar to the method proposed by~\citet{ref:clearyj} but based on the parametric equation of a line, $\boldsymbol{u} + \boldsymbol{v}t$ which makes implementation in 3D simple. Since its develpment, many improvements and extensions have been made for alternate geometries such as rhombic dodecahedra~\citep{ref:hel}

GPU RAYTRACING

@@

@@

Conceptually, code run on the device is split into block on a grid, each block then runs multiple concurrent computations, each on a thread. In hardware, threads execute on 32-128 streaming multiprocessors depending on the GPU architecture.

>>>>

\section{CT Dose Estimation}
As discussed in Chapter 1, the dose a patient receives from both diagnostic CT procedures and radiation therapy is of great concern to the medical community [CITE]. However, dose quantification is not currently patient specific. Measuring the dose in a patient directly is impossible, intead the air kerma (measured in Gy), which measures the x-ray beam intensity, and can be measured directly [CITE]. As the x-ray beam passes through the patient, it deposits energy. The energy deposition per unit mass is the absorbed dose (measured in Gy) which is then weighted by the radiation type to give the equivalent dose (measured in Sv). The equivalent dose quantifies the biological detriment to the patient from the procedure. The effective dose (measured in Sv) accounts for the radiosensitivity of each organ and is the most meaningful measure of risk to the patient. Any two procedures that result in the same effective dose to the patient carry the same acute and long-term risks. Any procedure that gives an effective dose to a patient has the same risk as any procedure that would give the same whole body dose.

@@

INSERT MORE EXAMPLES.
\citet{ref:lauk} found that severly obese patients received twice as much dose from CT procedures as others on average. Some patients received as much as five times the typical dose.

@@

Nearly all transport solutions fall into one of two categories: Monte Carlo and deterministic. Monte Carlo codes are generally known for their accuracy; they are able to sample the problem and attain arbitrarily high precision given sufficient runtime. They also take advantage of combinatorial geometry allowing a user to produce geometrically complex 3D structures. Deterministic methods are generally much faster than their Monte Carlo counterparts but do not benefit from excess runtime. They also consume vastly more resources, particularly random-access memory (RAM). However, in regularly structures meshes, deterministic methods can greatly outperform Monte Carlo. A large number of spatial regions slows down the particle transport in Monte Carlo and increases the RAM required to store the numerous voxels.

Codes such as EGS4, its improved versions EGSnrc, and EGS5 \citep{ref:nelsonw}, and PENELOPE \citep{ref:salvatf} are early Monte Carlo that have shown very good comparison to experimental results. However, these codes were developed primarily for shower simulations in high energy physics.

Monte Carlo codes such as TOPAS [CITE], Geant4 \citep{ref:agostinellis}, and MCNP6 [CITE] are the current gold standard in computational dosimetry \citep{ref:jiax}. This work compares results to those obtained from MCNP6. MCNP6 is a Monte Carlo code developed for general purpose particle transport and is well validated. MCNP6 was selected for comparison because it is so well validated and the input files can be easily generated procedurally. Codes whose input such as Geant4 do not lend themselves easily to procedural code generation are difficult to benchmark precisely against for arbitrary computational phantoms.

The alternative to Monte Carlo is deterministic solutions. The best known solution techniques include the discrete ordinate method, the finite element method, and the method of characteristics. The discrete ordinate method is employed by DOCTORS and the methodology is covered independently in Section~\ref{sec:discordlit}. Other deterministic methods are summarized here.

The method of characteristics solves the characteristic form of the Boltzmann transport equation which is defined along a single direction (called the characteristic). Along that direction, the 3-dimensional gradient operator found in the full Boltzmann equation (see Sec.~\ref{sec:lbe}) becomes a 1-dimensional derivitive. This reduction gives rise to an analytical solution along the characteristic [CITE]. Sufficient characteristic solutions solved simulateously gives the solution over the entire problem domain. Currently, though, MOC has not been extended to full 3D, instead, hybrid methods such as the NEM-$S_N$ technique employed by DeCART have been used. DeCART uses 2D MOC to solve planar slices of reactor transport problems and links the slices using diffusion~\citep{ref:hursinm}. Another MOC implementation is the TRANSMED code which was used to compute external beam therapy dose profiles accurately~\citep{ref:williamsm}. Method of characteristic solutions have been extended to the time dependent domain by \citet{ref:hoffmana}.

However the energy dependent flux is computed, it must afterward be converted to dose. Monte Carlo codes have been used in the past to compute flux-to-dose conversion factors for a reference person~\citep{ref:icrp116}. The values provided by~\citet{ref:icrp116} are energy dependent conversion factors, $H(E)$ that convert flux to effective dose for a particular orientation. The dose, $D$ is then computed
\begin{equation}
D = \int_0^\infty \varphi(E) H(E) dE
\end{equation}
where $\varphi$ is the scalar flux. The weakness of this method is that even though the flux may be computed to be patient specific, the weighting factors are evaluated using a standard reference man phantom.

Alternatively, codes can compute the dose deposition directly during the transport solution. The simplest example of this is in Monte Carlo codes which tally the energy deposited by each collision rather than track length in voxels which estimates the absorbed dose. Since the particle type that caused the scatter is always, known, the equivalent dose is simple to compute. However, in order to compute the effective dose, which is organ weighted, the organ to which the voxel of interest belongs must be known. In general, this is a very difficult problem with no fully automated solutions without significant additional information about human anatomy and the scan procedure employed. Therefore, this dose estimation technique is rarely employed in practice.

@@

The reconstruction populates the CT mesh with CT numers, also known as Hounsfield units. These values are related to the attenuation coefficient of the material represented by the voxel. Traditionally, a CT number of zero corresponds to the attenuation of water while -1000 corresponds to dry air [CITE] though some variations do exist \citep{ref:plessisf} \citep{ref:sawc}.

@@

\section{Discrete Ordinates}\label{sec:discordlit}
The dsicrete ordinates method dates back to [CITE] but remains one of the most prominent solution modalities for radiation transport in use today. A comprehensive review of discrete ordinate methods is given by \citet{ref:lewise}. Additional references include [CITES].

The oldest major discrete ordinate method implementation is DORT which was superceded by its 3D counterpart TORT~\citep{ref:rhoadesw}. Since TORT, new discrete ordinate method implementations have laregly been extensions allowing more advanced computation or updates to modernize the code.

@@

To overcome ray effect artifacts (discussed more thouroughly in Sec.~\ref{sec:uncol}), discrete ordinate solutions often employ a raytracing algorithm to compute the uncollided flux. The uncollided flux is then computed using the full discrete ordinate method. Raytracing algorithms have been developed for numerous discret ordinate codes include RAY3D~\citep{ref:yingz} and ATTILLA~\citep{ref:wareingt} which utilizes unstructured tetrahedral meshes.

INSERT KBA DISCUSSION

@@

Many raytracing algorithms are based on the Bresenham line raserizing algorithm which was designed for graph plotting hardware~citep{ref:bresenhamj}. That algorithm was intended to produce lines acceptable for human interpretation and no guarantee is made that all pixels along the line will be identified making this class of algorithms inappropriate for dosimetric raytracing since attenuation through some voxels would be skipped. Some extensions have been added to it that guarantee passage through all voxels though~\citep{ref:liuy}.

\citet{ref:clearyj} proposed a voxel traversal algorithm in which he initially identifies the voxel containing the starting point of the ray, and transport it to the first surface of its bounding voxel it would cross. This is done by computing the distance traveled to cross each enclosing surface. The surface with the least distance is the first surface crossed. Therefore, the particle is advanced to the intersection point and considered inside the next cell, having crossed the surface. This process is repeated until a termination criteria is met.

This work uses the raytracing algorithm proposed by~\citet{ref:wooa} which is very similar to the method proposed by~\citet{ref:clearyj} but based on the parametric equation of a line, $\boldsymbol{u} + \boldsymbol{v}t$ which makes implementation in 3D simple. Since its develpment, many improvements and extensions have been made for alternate geometries such as rhombic dodecahedra~\citep{ref:hel} for volume rendering but the core algorithm remains state of the art for floating-point arithmetic. On systems that are highly efficient at computing integer arithmetic, alternatives exist which are faster~\citep{ref:liuy} but they do not outperform on hardware optimized for floating-point operations.

INSERT SECTION ON GPU RAYTRACING

@@

A graphics processing unit (GPU) is similar to a central processing unit (CPU)in that it executes instructions to process data. However, it differs in that a CPU may have up up to 22 physical cores, each of which can operate independently, a GPU has up to thousands of cores, but they cannot operate independently. Instead, cores execute simulaneously in groups called warps. All cores in a warp execute the same instruction at the same time as all others.

@@

Conceptually, code run on the device is split into block on a grid, each block then runs multiple concurrent computations, each on a thread. In hardware, threads execute in a warp (32 to 128 streaming multiprocessors depending on the GPU architecture).

### 2017-06-08 docprep/chapconc.tex
Updated thesis text

I concluded many things.

@@

\subsection{Partial Acceptance Criteria}
Allow it in the raytracer.

\subsection{Multileaf Collimation}
Collimators!

>>>>

The raytracer is very fast compared to the discrete ordinate solution which computes the collided flux. The uncollided flux quantifies the source distribution from the medical system. This uniquely enables discrete ordinate solutions to rapidly characterize complex systems. Multiple cone beams and fan beams are possible as well as many other, more complex beam shapes.

Anisotropy cannot be ignored.

@@

\subsection{Partial Acceptance Criteria}
In the current version of doctors, the raytracer is very accurate with respect to MCNP6 in predicting the uncollided flux, except for along the periphery of the beam. Voxels are currently either completely inside the beam or completely outside of it as determined by its isocenter. A more sophisticated technique whereby a voxel cah be partially accepted in the beam would remove these artifacts.

\subsection{Multileaf Collimation}
Some medical systems currently utilize a multileaf collimator [CITE] to deliver dose to a specific organ or region of the body. As the
